Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-02-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: add more text describing clustered bar chart results

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    12      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
Working-set size (bytes)   & $56n^2$ &  $72n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. $n$ is the problem size -- the length of one side of the
  square domain.}
\label{TAB_kernel_details}
\end{table}

@@

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1.1~MB and 4.7~MB, respectively and therefore
both of these should fit comfortably within L3 cache. Despite this,
the performance achieved with these domain sizes falls short of the L3
bandwidth limit.  Although this kernel is one of several within the
time-stepping loop of NEMOLite2D, the working-set size of the latter
for the $128^2$ domain is just 2.4~MB which again is well within the
capacity of L3.

@@

In order to investigate why the performance of the Continuity kernel
is falling well short of the maximum predicted by the roofline model
we have experimented with the level of SIMD
vectorisation. Figure~\ref{FIG_cont_vec} shows the performance of the
Continuity kernel compiled with differing levels of SIMD vectorisation
and after code modifications intended to improve vectorisation.
Compiling without auto-vectorisation (``-no-vec''), the kernel
achieves a performance that is largely limited by the bandwidth to
main memory. Note that there is very little opportunity for temporal
cache re-use in the Continuity kernel -- only three of the twelve
distinct memory references are to previous rows of the array. The size
of this effect can be seen in the small drop in performance (black
bars) in going from the $256^2$ to the $512^2$ domain as at this point
the working-set size spills out of L3 cache.

Figure~\ref{FIG_kernel_perf} shows the performance of
the Continuity and u-Momentum kernels when compiled without
vectorisation (``-no-vec''), with SSE4 (Streaming SIMD Extensions)
vectorisation (``-axSSE4.2'') and with AVX (``-xHost''). SSE supports
vector lengths of 128 bits (two double-precision words) and AVX 512
bits (four double-precision words). For the 64, 128 and 256 problem
sizes, changing from no vectorisation to SSE increases the performance
of the Continuity kernel by up to roughly 60\%; some way short of the
ideal speed-up of 100\% that would be obtained with perfect
SSE vectorisation of a compute-bound kernel. Unsurprisingly then, moving
from SSE to AVX vectors does not give any performance improvement and
in fact, is slightly detrimental for most of the problem sizes.

@@

\begin{figure}
  \centering
  \includegraphics[width=130mm]{continuity_opt}
  \caption{Performance of the Continuity kernel with optimisations
    intended to aid SIMD vectorisation. Results are for v.16 of the
    Intel compiler on a Xeon E5-1620 v.2 CPU.}
  \label{FIG_cont_vec}
\end{figure}

>>>>

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
                             & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
WSS of time-stepping (bytes) & $104n^2$ & \multicolumn{2}{c|}{$180n^2$} \\ 
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    12      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.128     &   0.42     \\
WSS (bytes)                & $56n^2$ &  $72n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. WSS is the Working Set Size and $n$ is the problem size --
  the length of one side of the square domain.}
\label{TAB_kernel_details}
\end{table}

@@

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1.1~MB and 4.7~MB, respectively and
therefore both of these should fit comfortably within L3
cache. Despite this, the performance achieved with these domain sizes
falls short of the L3 bandwidth limit.  However, the Continuity kernel is just
one of several within the time-stepping loop of NEMOLite2D; the loop
as a whole has a working-set size of $180\times n^{2}$ bytes. For the
$32^2$ domain this is 180~KB, for $52^2$ it is 475~KB and for the
$128^2$ domain it is 2.9~MB which is well within the capacity of L3.

@@

In order to investigate why the performance of the Continuity kernel
is falling well short of the maximum predicted by the roofline model
we have experimented with the level of SIMD
vectorisation. Figure~\ref{FIG_cont_vec} shows the performance of the
Continuity kernel compiled with differing levels of SIMD vectorisation
and after code modifications intended to improve vectorisation.
Compiling without auto-vectorisation (compiler flag ``-no-vec''), the
kernel achieves a performance that appears to be largely limited by
the bandwidth to main memory. Note that there is very little
opportunity for temporal cache re-use in the Continuity kernel -- only
three of the twelve distinct memory references are to previous rows of
the array. The size of this effect can be seen in the small drop in
performance (black bars) in going from the $256^2$ to the $512^2$
domain as at this point the working-set size of this kernel spills out
of L3 cache.

\begin{figure}
  \centering
  \includegraphics[width=130mm]{continuity_opt}
  \caption{Performance of the Continuity kernel with optimisations
    intended to aid SIMD vectorisation. Results are for v.16 of the
    Intel compiler on a Xeon E5-1620 v.2 CPU. Optimisations are
    cumulative for results with a given SIMD instruction set.}
  \label{FIG_cont_vec}
\end{figure}

The E5-1620 CPU supports two different SIMD instruction sets;
Streaming SIMD Extensions (SSE; flag ``-axSSE4.2'') and Advanced
Vector Extensions (AVX; flag ``-xHost''). SSE supports vector lengths
of 128 bits (two double-precision words) and AVX 256 bits (four
double-precision words). For the problem sizes that largely fit within
L3 or L2 cache, switching on the use of SSE instructions provides a
speed-up of between 1.5 and 1.8 - the factor increasing with problem
size until the WSS of the time-stepping loop begins to fall out of L3
cache. It is therefore likely that the reason this factor is below the
ideal value of 2.0 is the overhead associated with constructing the
loop and handling any ``peel'' or ``remainder'' loops. As the problem
size increases, this overhead (as a fraction of the vectorisable
workload) decreases.

Figure~\ref{FIG_kernel_perf} shows the performance of the Continuity
and u-Momentum kernels when compiled without vectorisation
(``-no-vec''), with SSE vectorisation (``-axSSE4.2'') and with AVX
(``-xHost''). SSE supports vector lengths of 128 bits (two
double-precision words) and AVX 512 bits (four double-precision
words). For the 64, 128 and 256 problem sizes, changing from no
vectorisation to SSE increases the performance of the Continuity
kernel by up to roughly 60\%; some way short of the ideal speed-up of
100\% that would be obtained with perfect SSE vectorisation of a
compute-bound kernel. Unsurprisingly then, moving from SSE to AVX
vectors does not give any performance improvement and in fact, is
slightly detrimental for most of the problem sizes.

@@
